The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2018-09-27 article.tex
Updates from ShareLaTeX

\gio{I've taken the freedom to change your proposed structure. Feel free to change whatever you don't like it. I might be not completely accurate on the biological background.}
\richel{It's just fine :-)}

Current phylogenetic tools assume that only a single speciation event can occur at the same time.
While this assumption is useful to construct a wide variety of successful models \richel{@gio: citation here},
they disallow for environmental changes that trigger speciation on a large scale,
for example, the cichlid fish diversification in the 
African Great Lakes: Malawi, Tanganyika and Victoria \richel{@gio: citation here}.

@@

\gio{If I described the process in the same way you report in the example I would probably end up writing the same things that we say a few lines below, where we describe the parameters. Don't you think?}
\richel{You described the model in the Methods. I moved it to here}
In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the usual per-species speciation and extinction rates. 
The model also introduces two new parameters: 
$\nu$ is the total rate for an environmental change to trigger, 
which leads to a potentially multiple speciation event 
\richel{How exactly? When the environment change is triggered, is the speciation rate $q$ or $\lambda + q$? What ends the change?}.
If such event triggers each species present at that moment in time can undergo a speciation event with probability $q$.

@@

We simulate protracted birth-death trees, using the MBD package (\cite{pbd}) in the R programming language (\cite{r}).
The first tree has a random number generator seed of 1, which is incremented by 1 for each simulated tree.
For each combination of $\lambda, \mu, \nu$ and $q$, 
we generate species trees with a crown age of 15 million years 
\gio{are we sure we wanna try 15 million years. so far i've been trying only with 10 million years, which most of the time is working really well}
\richel{I suggest yes. Most papers associated with Rampal use 15 million years, including my raket paper. Just use lower speciation rates}.
Only trees with the desired number of good taxa are kept.

@@

\richel{must rewrite, use pirouette as a starting point}
From an alignment, we run a Bayesian analysis and 
create a posterior distribution of trees and parameters
using the \verb;pirouette; (\cite{pirouette}) package
that sets the input parameters similar to BEAUti 2 and then runs BEAST2. 
For our site model, we assume either a Jukes-Cantor or GTR nucleotide substitution model.
The Jukes-Cantor model is the correct one, as it is used for simulating that alignment,
where the GTR model is the site model that is picked as a default by most users.
For our clock model, we assume either a strict or relaxed log-normal 
clock model. 
Also here, the strict clock model is the correct one, as it is used for simulating the alignment,
but the relaxed log-normal clock model is the one most commonly used.
We set the BD model as a tree prior, 
as gauging the effect of this incorrect assumption is the goal of this study. 
We assume an MRCA prior with a tight normal distribution
around the crown age, by choosing the crown age as mean, and a standard deviation 
of $0.5 \cdot 10^{-3}$ time units,
resulting in 95\% of the crown ages inferred have the same resolution (of $10^{-3}$ time 
units) as the alignment. 
We ran the MCMC chain to generate 1111 states,
of which we remove the first 10\% (also called the 'burn-in'). 
Of the remaining
1000 MCMC states, the effective sample size (ESS) of the posterior 
must at least be 200
for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
the number of samples or decreasing the autocorrelation between samples. 
If the ESS is less than 200, we decrease autocorrelation by doubling 
the MCMC sampling interval of that simulation, until the ESS exceeds 200.

@@

The input trees generated with a \richel{@gio: parameter that is set to reduce the MBD model to BD} 
allow us to measure the noise of the experiment.

@@

% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MEE style
\bibliographystyle{mee}
\bibliography{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{p{0.05\textwidth} p{0.6\textwidth} p{0.2\textwidth}}
    \hline
                          & Description & Values \\
    \hline
    \hline
    $\lambda$             & \richel{@gio: MBD params here} & 0.1, 0.3, 1.0, $10^9$ \\
    $\lambda$             & Speciation rate & 0.1, 0.3, 1.0, $10^9$ \\
    $\mu$                 & Extinction rate & 0.0, 0.1, 0.2 \\
    \hline
    $n$                   & Number of good taxa & 50, 100, 200 \\
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M_s$                 & Sampling method & S, L, R \\
    $M_c$                 & Clock model & S, RLN \\
    $M_t$                 & Site model & JC69, GTR \\
    $r$                   & Mutation rate & $\frac{1}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1, 2, etc. \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the simulation parameters. Above the horizontal line is 
    the biological parameter set. 
    The RNG seed $R_i$ is 1 for the first simulation, 2 for the next,
    and so on.  
    The clock models are abbreviated as 'S' for a strict and 'RLN' for a relaxed log-normal model.
    The site models are abbreviated as 'JC69' for Jukes-Cantor (\cite{jc69}) and 'GTR' for the generalized 
    time-reversible model (\cite{gtr}).
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\richel{put this section here, as the journal does not request for this}
RSE \richel{@gio: I assume this this is true?} conceived the idea for this experiment.
GL created and tested the MBD package.
RJCB created and tested the experiment.
GL and RJCB wrote the first draft of the manuscript. 
RSE contributed substantially to revisions.

>>>>

Current phylogenetic tools assume that only a single speciation event can occur at the same time.
While this assumption is useful to construct a wide variety of successful models \richel{@gio: citation here} \gio{@richel: basically all the models we know are based on this assumpion: DDD, PBD, BISSE, MUSSE, SECSSE, any other SSE, DAISIE etc. etc. It's a very very general feature. Maybe being specific could lead the reader to consider things that are, in the end, not essential to the story we want to tell here. Do you still think we need it?},
they disallow for environmental changes that trigger speciation on a large scale,
for example, the cichlid fish diversification in the 
African Great Lakes: Malawi, Tanganyika and Victoria \richel{@gio: citation here}.

@@

\gio{If I described the process in the same way you report in the example I would probably end up writing the same things that we say a few lines below, where we describe the parameters. Don't you think?}
\richel{You described the model in the Methods. I moved it to here}
In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the usual per-species speciation and extinction rates. 
The model also introduces two new parameters: 
$\nu$ is the total rate for an environmental change to trigger, 
which leads to a potentially multiple speciation event 
\richel{How exactly? When the environment change is triggered, is the speciation rate $q$ or $\lambda + q$? What ends the change?}.
\gio{If the environmental change triggers only $q$ is taken in consideration. $\lambda + q$ is not even a possibility, as $\lambda$ is a rate while $q$ is a probability. Do you think we need to be clearer on that?}
If such event triggers each species present at that moment in time can undergo a speciation event with probability $q$.

@@

We simulate protracted birth-death trees, using the MBD package (\cite{pbd}) in the R programming language (\cite{r}).
The first tree has a random number generator seed of 1, which is incremented by 1 for each simulated tree.
For each combination of $\lambda, \mu, \nu$ and $q$, 
we generate species trees with a crown age of 15 million years 
\gio{are we sure we wanna try 15 million years. so far i've been trying only with 10 million years, which most of the time is working really well}
\richel{I suggest yes. Most papers associated with Rampal use 15 million years, including my raket paper. Just use lower speciation rates}
\gio{In general it is ok for me. Keep in mind, though, that allowing multiple speciations may lead to an explosion in the number of species. Increasing the time by a factor of n usually means increasing the expected number of species at the present by a factor proportional to $e^n$}.
Only trees with the desired number of good taxa are kept.

@@

\richel{must rewrite, use pirouette as a starting point}
From an alignment, we run a Bayesian analysis and 
create a posterior distribution of trees and parameters
using the \verb;pirouette; (\cite{pirouette}) package
that sets the input parameters similar to BEAUti 2 and then runs BEAST2. 
For our site model, we assume either a Jukes-Cantor or GTR nucleotide substitution model.
The Jukes-Cantor model is the correct one, as it is used for simulating that alignment,
where the GTR model is the site model that is picked as a default by most users.
For our clock model, we assume either a strict or relaxed log-normal 
clock model. 
Also here, the strict clock model is the correct one, as it is used for simulating the alignment,
but the relaxed log-normal clock model is the one most commonly used.
We set the BD model as a tree prior, 
as gauging the effect of this incorrect assumption is the goal of this study. 
We assume an MRCA prior with a tight normal distribution
around the crown age, by choosing the crown age as mean, and a standard deviation 
of $0.5 \cdot 10^{-3}$ time units,
resulting in 95\% of the crown ages inferred have the same resolution (of $10^{-3}$ time 
units) as the alignment. 
We ran the MCMC chain to generate 1111 states,
of which we remove the first 10\% (also called the 'burn-in'). 
Of the remaining
1000 MCMC states, the effective sample size (ESS) \gio{@richel: could it be nicer, when citing the "effective sample size" for the first time (in order to use the acronym thereafter), to put all in capital letters "Effective Sample Size"} of the posterior 
must at least be 200
for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
the number of samples or decreasing the autocorrelation between samples. 
If the ESS is less than 200, we decrease autocorrelation by doubling 
the MCMC sampling interval of that simulation, until the ESS exceeds 200.

@@

The input trees generated with a \richel{@gio: parameter that is set to reduce the MBD model to BD}\gio{@richel: setting $nu = 0$ denies any chance for multiple speciations to occur. If so, $q$ setting becomes unimportant. As a convention is possible to set $q=0$} 
allow us to measure the noise of the experiment.

@@

% Bibliography
\gio{bibliography is missing. The only bib file present does not correspond to the bibliography showed in the pdf file.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MEE style
\bibliographystyle{mee}
\bibliography{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{p{0.05\textwidth} p{0.6\textwidth} p{0.2\textwidth}}
    \hline
                          & Description & Values \\
    \hline
    \hline
    $\lambda$             & \gio{@richel: I insert some params that can make sense. Let's keep en eye on this anyways} & 0.1, 0.3, 1.0, $10^9$ \\
    $\lambda$             & Per-species speciation rate & 0.2 \\
    $\mu$                 & Per-species extinction rate & 0.0, 0.15 \\
    $\nu$                 & Multiple speciation trigger rate & 1.5, 2.0, 2.5 \\
    $q$                   & Per-species probability of multiple speciation  & 0.1, 0.15 \\
    \hline
    \hline
    $n$                   & Number of good taxa & 50, 100, 200 \\
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M_s$                 & Sampling method & S, L, R \\
    $M_c$                 & Clock model & S, RLN \\
    $M_t$                 & Site model & JC69, GTR \\
    $r$                   & Mutation rate & $\frac{1}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1, 2, etc. \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the simulation parameters. Above the horizontal line is 
    the biological parameter set. 
    The RNG seed $R_i$ is 1 for the first simulation, 2 for the next,
    and so on.  
    The clock models are abbreviated as 'S' for a strict and 'RLN' for a relaxed log-normal model.
    The site models are abbreviated as 'JC69' for Jukes-Cantor (\cite{jc69}) and 'GTR' for the generalized 
    time-reversible model (\cite{gtr}).
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\richel{put this section here, as the journal does not request for this}
RSE \richel{@gio: I assume this this is true?} \gio{@richel: it is true} conceived the idea for this experiment.
GL created and tested the MBD package.
RJCB created and tested the experiment.
GL and RJCB wrote the first draft of the manuscript. 
RSE contributed substantially to revisions.
